The Two Dualities of Computation: Negative and Fractional Types

### 2012-02-28 rational.tex
fixed some latex issues and small edits

\usepackage{graphicx}
\usepackage{longtable}
\usepackage{comment}
\usepackage{amsmath}
\usepackage{mdwlist}
\usepackage{txfonts}
\usepackage{xspace}
\usepackage{amstext}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{proof}
\usepackage{multicol}
\usepackage[nodayofweek]{datetime}
\usepackage{etex}
\usepackage[all, cmtip]{xy}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{multicol}
\usepackage{bm}
\usepackage{cmll}

@@

%subcode-inline{bnf-inline} name langRev
%! swap+ = \mathit{swap}^+
%! swap* = \mathit{swap}^*
%! dagger =  ^{\dagger}
%! assocl+ = \mathit{assocl}^+
%! assocr+ = \mathit{assocr}^+
%! assocl* = \mathit{assocl}^*
%! assocr* = \mathit{assocr}^*
%! identr* = \mathit{uniti}
%! identl* = \mathit{unite}
%! dist = \mathit{distrib}
%! factor = \mathit{factor}
%! eta+ = \eta^+
%! eps+ = \epsilon^+
%! eta* = \eta^*
%! ^^^ = ^{-1}
%! eps* = \epsilon^*
%! (o) = \fatsemi
%! (;) = \fatsemi
%! (*) = \times
%! (+) = +

@@

We show a deep symmetry between functions and delimited continuations,
values and continuations that arises in {{langRev}} in a manner that
is reminiscent of Filinksi's Symmetric \lcal
~\cite{Filinski:1989:DCI:648332.755574}. The symmetry arises by
extending {{langRev}} with a notion of additive duality over the
monoid {{(+, 0)}} by including {{eta}} and {{eps}} operators of
Compact Closed Categories. The resulting dual types, which we denote
{{-b}}, have a time traveling ``backward information flow''
interpretation and allow for the encoding of higher-order function and
iteration via the construction of {{trace}} operators, thereby making
the extended language {{langRevEE}} a Turing-complete reversible
programming language with higher-order functions and first-class
delimited continuations.

@@

\begin{proposition}
\label{prop:groupoid}
{{langRev}} is a dagger symmetric monoidal category. 
\end{proposition}

@@

@@

@@

Note that we will freely write {{b^^^}} as {{1/b}} and {{a*b^^^}} as
{{a/b}}.

@@

\begin{proposition}[Groupoid]
\label{prop:groupoid}
{{langRevEE}} is a groupoid. 
\end{proposition}

\begin{proposition}
\label{prop:groupoid}
{{langRev}} is a Dagger Symmetric Monoidal Closed Category.
\end{proposition}

@@

Square root and imaginary types have also appeared in the literature: we
relegate the connections to Sec.~\ref{related} and proceed with a simple
explanation. We have so far extended the set of types to be the rational
numbers. Now we will push this and extend the set of types to algebraic
numbers. In other words, we will allow datatypes defined by arbitrary
polynomials and allow the roots of such polynomials to be types.

>>>>

\usepackage{graphicx}
\usepackage{longtable}
\usepackage{comment}
\usepackage{amsmath}
\usepackage{mdwlist}
\usepackage{txfonts}
\usepackage{xspace}
\usepackage{amstext}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{proof}
\usepackage{multicol}
\usepackage[nodayofweek]{datetime}
\usepackage{etex}
\usepackage[all, cmtip]{xy}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{multicol}
\newcommand\hmmax{0} % default 3
\newcommand\bmmax{0} % default 4
\usepackage{bm}
\usepackage{cmll}

@@

%subcode-inline{bnf-inline} name langRev
%! swap+ = \mathit{swap}^+
%! swap* = \mathit{swap}^*
%! dagger =  ^{\dagger}
%! assocl+ = \mathit{assocl}^+
%! assocr+ = \mathit{assocr}^+
%! assocl* = \mathit{assocl}^*
%! assocr* = \mathit{assocr}^*
%! identr* = \mathit{uniti}
%! identl* = \mathit{unite}
%! dist = \mathit{distrib}
%! factor = \mathit{factor}
%! eta+ = \eta^+
%! eps+ = \epsilon^+
%! eta* = \eta^{\times}
%! eps* = \epsilon^{\times}
%! trace+ = trace^+
%! trace* = trace^{\times}
%! ^^^ = ^{-1}
%! (o) = \fatsemi
%! (;) = \fatsemi
%! (*) = \times
%! (+) = +

@@

We show a deep symmetry between functions and delimited continuations,
values and continuations that arises in {{langRev}} in a manner that
is reminiscent of Filinksi's Symmetric \lcal
~\cite{Filinski:1989:DCI:648332.755574}. The symmetry arises by
extending {{langRev}} with a notion of additive duality over the
monoid {{(+, 0)}} by including {{eta+}} and {{eps+}} operators of
Compact Closed Categories. The resulting dual types, which we denote
{{-b}}, have a time traveling ``backward information flow''
interpretation and allow for the encoding of higher-order function and
iteration via the construction of {{trace}} operators, thereby making
the extended language {{langRevEE}} a Turing-complete reversible
programming language with higher-order functions and first-class
delimited continuations.

@@

\begin{proposition}
\label{prop:category}
{{langRev}} is a dagger symmetric monoidal category. 
\end{proposition}

@@

Multiplicative duality is added to {{langRev}} by adding types
{{b^^^}}, values {{v^^^}} and isomorphisms {{eta*}} and {{eps*}}.
Note that we will freely write {{b^^^}} as {{1/b}} and {{a*b^^^}} as
{{a/b}}.

@@

\noindent
These have the type judgments

@@

@@

\begin{proposition}[Logical Reversibility]
\label{prop:logrevEE}
{{c v |--> v'}} iff {{c{dagger} v' |--> v}}.
\end{proposition}

\begin{proposition}[Groupoid]
\label{prop:groupoidEE}
{{langRevEE}} is a groupoid. 
\end{proposition}

\begin{proposition}
\label{prop:categoryEE}
{{langRev}} is a Dagger Symmetric Monoidal Closed Category.
\end{proposition}

@@

Square root and imaginary types have also appeared in the literature: we
relegate the connections to Sec.~\ref{sec:related} and proceed with a simple
explanation. We have so far extended the set of types to be the rational
numbers. Now we will push this and extend the set of types to algebraic
numbers. In other words, we will allow datatypes defined by arbitrary
polynomials and allow the roots of such polynomials to be types.
